# The Coalescent

After introducing and defining genealogies in the previous chapter, John Wakeley talks about the statistical properties of genealogies under the coalescent framework. To demonstrate the coalescent framework, he uses the classical population genetics models:

a.  Wright Fisher Model
b.  Moran Model

and shows these models converging to the Coalescent in the limit.

<br>

Throughout this chapter, there is no reference to variation in sequences with the following two reasons:

1.  A sample without any variation still supports a genealogy, because of the way DNA replication takes place

2.  All variation is assumed to be selectively neutral (any selection is dealt later in the book) and so there is no population advantage with respect to any particular variation.

## Basic Population Genetic Models

### Wright Fisher Model {.unnumbered}

-   Perfectly **non-overlapping** generations
-   Population size is constant and finite at all times
-   Each generation all current individuals die and are replaced by their offsprings. This is a random process with the possibility that a current individual has no offsprings in the next generation. For any individual in the next generation, any individual in the current generation has an equal chance of being a parent.

Let's consider a simple example of a population with two alleles **A** and **B** following the WF model. Let the population size be $N$ and $K_t$ be the no. of **A** alleles in generation $t$ with $K_0 = i$. Then the probability that the $K_1 = j$ is given by: $$\mathbb{P}_{ij} = {N \choose j}\left(\frac{i}{N}\right)^j\left(1-\frac{i}{N}\right)^{N-j}$$ We also have that the expected number and variance of **A** alleles in generation $1$ as $$\begin{align*} & \mathbb{E}(K_1) = N\left(\frac{i}{N}\right) = i \\ & \mathbb{V}ar(K_1) = N \left(\frac{i}{N}\right) \left(1 - \frac{i}{N}\right) = i\left(1-\frac{i}{N}\right)  \end{align*}$$

The expected number of **A** alleles remains unchanged over time.

### Moran Model {.unnumbered}

-   Perfectly **Overlapping** Generations
-   Population size is again constant and finite over time
-   At the end of every generation, two current individuals are chosen. One individual survives and gives rise to an offspring where as the other perishes.

Consider the same example with the population following the Moran Model instead of the WF Model. Again let $K_t$ be the no. of **A** allele in generation $t$. Then with $K_0=i$, probability that $K_1=j$ is given by: $$ \mathbb{P}_{ij} = \begin{cases} p(1-p) & j=i+1 \\ (1-p)p & j=i-1 \\ p^2 + (1-p)^2 & j=i \\ 0 & \text{otherwise} \end{cases}    $$

where ($p=\frac{i}{N}$). Again the expected number and variance of no. of **A** alleles is given by: $$\begin{align*}  \mathbb{E}(K_1) =& p(1-p)(i+1) + p(1-p)(i-1) + (p^2+(1-p)^2)i \\ =& i \\  \mathbb{V}ar(K_1) =& p(1-p)(i+1)^2 + p(1-p)(i-1)^2 + (p^2+(1-p)^2)i^2  -i^2 \\ =& 2p(1-p)  \end{align*}$$

There are two steps to cross-reference any heading:

1.  Label the heading: `# Hello world {#nice-label}`.
    -   Leave the label off if you like the automated heading generated based on your heading title: for example, `# Hello world` = `# Hello world {#hello-world}`.
    -   To label an un-numbered heading, use: `# Hello world {-#nice-label}` or `{# Hello world .unnumbered}`.
2.  Next, reference the labeled heading anywhere in the text using `\@ref(nice-label)`; for example, please see Chapter.
    -   If you prefer text as the link instead of a numbered reference use: [any text you want can go here].

## Captioned figures and tables

Figures and tables *with captions* can also be cross-referenced from elsewhere in your book using `\@ref(fig:chunk-label)` and `\@ref(tab:chunk-label)`, respectively.

See Figure \@ref(fig:nice-fig).

```{r nice-fig, fig.cap='Here is a nice figure!', out.width='80%', fig.asp=.75, fig.align='center', fig.alt='Plot with connected points showing that vapor pressure of mercury increases exponentially as temperature increases.'}
par(mar = c(4, 4, .1, .1))
plot(pressure, type = 'b', pch = 19)
```

Don't miss Table \@ref(tab:nice-tab).

```{r nice-tab, tidy=FALSE}
knitr::kable(
  head(pressure, 10), caption = 'Here is a nice table!',
  booktabs = TRUE
)
```
